---
title: "Introduction to hmagmaR"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Introduction to hmagmaR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

`hmagmaR` links non-coding GWAS SNPs to their putative target genes by combining
chromatin interaction data (e.g. PLAC-seq or Hi-C) with regulatory region annotations
(e.g. promoters, enhancers, H3K27ac peaks). The output is a MAGMA-compatible gene
annotation file, which can then be used for gene-level association testing.

The typical workflow has two stages:

1. **`AnnotationFileHmagma()`** (or its down-sampled variant, `SampledDownAnnotation()`)
   builds the gene annotation file by overlapping chromatin interactions with regulatory
   regions and SNPs.
2. **`GeneLevelAnalysis_hmagma()`** runs MAGMA itself on that annotation file against a
   GWAS summary statistics file, using an external MAGMA binary and 1000 Genomes
   reference panel.

This vignette demonstrates stage 1 using a small example dataset bundled with the
package. Stage 2 requires external software and reference files not included with
`hmagmaR`, so it is shown for reference only and not executed here.

```{r setup}
library(hmagmaR)
```

## Example data

`hmagmaR` ships with a small example dataset, `test_dataset.microglia.hg19`, subsampled
from microglia PLAC-seq interactions (Nott et al., 2019), microglia promoter/enhancer
regions, 1000 Genomes SNPs, and a GWAS summary statistics table, so the workflow below
runs in seconds rather than requiring genome-wide reference data.

```{r load-data}
data("test_dataset.microglia.hg19")

hic[1:3, ]
promoterRegions[1:3, ]
enhancerRegions[1:3, ]
snps[1:3, ]
annotated_genes[1:3, ]
snpgeneexon[1:3, ]
```

- `hic`: chromatin interaction data with columns `chrom1`, `start1`, `end1`, `chrom2`,
  `start2`, `end2` (plus additional interaction-calling metadata).
- `promoterRegions` / `enhancerRegions`: regulatory regions in BED-like format
  (`chr`, `start`, `end`).
- `snps`: reference SNPs (e.g. from a `.bim` file), with columns `chr`, `Position`, `SNP`.
- `annotated_genes`: reference gene coordinates, with columns `chr`, `start`, `end`, `ensg`.
- `snpgeneexon`: SNPs falling within exons/promoters of each gene, with columns `rsid`,
  `ensg`.

## Building a gene annotation file

`AnnotationFileHmagma()` accepts either a single `regulatoryRegions` table, or separate
`promoterRegions` and `enhancerRegions` tables (in which case promoters anchor the
chromatin interactions, and enhancers filter which SNPs are retained). Here we use the
latter:

```{r run-annotation}
annotation_file <- tempfile("hmagmaR_example")

AnnotationFileHmagma(
  hic = hic,
  promoterRegions = promoterRegions,
  enhancerRegions = enhancerRegions,
  snps = snps,
  annotated_genes = annotated_genes,
  snpgeneexon = snpgeneexon,
  AnnotationFile = annotation_file
)
```

This writes a MAGMA-style annotation file (`<AnnotationFile>.transcript.annot`), with one
row per gene: the Ensembl gene ID, its genomic coordinates, and the tab-separated rsIDs
of all SNPs linked to it.

```{r show-output}
result <- readLines(paste0(annotation_file, ".transcript.annot"))
length(result)
head(result, 5)
```

With this small example dataset, `r length(result)` genes were annotated with at least
one interacting SNP. Because the bundled data is a random subsample of the full
reference tables rather than a curated locus, this number is illustrative only - a real
analysis using full genome-wide interaction, regulatory region, and SNP reference data
will typically annotate many more genes.

## Downstream: gene-level analysis with MAGMA

Once an annotation file has been generated, `GeneLevelAnalysis_hmagma()` calls MAGMA to
perform gene-level association testing against a GWAS summary statistics file. This step
requires:

- A local installation of [MAGMA](https://cncr.nl/research/magma/),
- 1000 Genomes reference files (`.bed`/`.bim`/`.fam`) for the appropriate ancestry,
- GWAS summary statistics with `SNP`, `P`, and `N` columns.

```{r gene-level-analysis, eval = FALSE}
GeneLevelAnalysis_hmagma(
  magma = "/path/to/magma",
  g1000 = "/path/to/g1000_eur",
  gwas = "/path/to/gwas_summary_stats.txt",
  AnnotationFile = paste0(annotation_file, ".transcript.annot"),
  output = "hmagmaR_example"
)
```

This produces MAGMA's standard gene-level output files (`.genes.out`, `.genes.raw`,
`.log`), which can be used for downstream gene set or pathway analyses.

## Sampling down (`SampledDownAnnotation()`)

When comparing annotation results across samples with different numbers of chromatin
interactions, `SampledDownAnnotation()` repeatedly subsamples the `hic` table down to a
fixed number of interactions (`loopNumber`), so that differences in results reflect
biology rather than differing input sizes. It otherwise follows the same pipeline as
`AnnotationFileHmagma()` with a single combined `regulatoryRegions` table, writing one
annotation file per resample (suffixed `.transcript.annot.1`, `.transcript.annot.2`,
etc.).

## Session information

```{r session-info}
sessionInfo()
```
